Nitric oxide feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2023-05-24 Manuscript/Manuscript.Rmd
MS: removed bold from figure refs

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences **(Figure 1---figure supplement 1)**.
To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis.
In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region **(Figure 1D and Figure 1---figure supplement 2)**.
*NOS* was also expressed in the region of the visual eyes (adult eyes) and the pigmented eyespots **(Figure 1---figure supplement 2)**.
The four apical organ cells, but not the eyes, were also labelled with a *NOS* reporter construct driving palmitoylated-tdTomato **(Figure 1E)**.
This reporter also revealed the axonal projections of these central NOS-expressing neurons.
The position and morphology of the four *NOS+* cells allowed us to identify the same four cells as four interneurons (INNOS) in our three-day-old whole-body *Platynereis* volume EM data [@williams2017; @veraszto2020] **(Figure 1B,C)**.
In the synaptic connectome, the INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets **(Figure 1C,F)**. The projections of INNOS cells are segregated into input and output compartments, with cPRCs inputs occurring in the dense neurosecretory plexus. INNOS output synapses form in the more ventral projection region **(Figure 1---figure supplement 3)**.
INRGW neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the prototroch ciliary band and the cholinergic MC ciliomotor neurons **(Figure 1C,F)** [@Veraszto2017].

@@

The expression of *NOS* in the INNOS interneurons in the cPRC circuit suggests that NO signalling may be involved in UV/violet-avoidance.
To test this, first we asked whether NO is produced during UV/violet stimulation of the larvae.
We injected the fluorescent NO-reporter DAF-FM into zygotes and imaged two-day-old larvae while exposing the region of cPRC cilia to 405 nm violet light. To mark cell outlines, we coinjected mRNA encoding a red-fluorescent reporter (RGECO), allowing the identification of the cPRCs.
Following light stimulation in the region of the ramified cilia of the cPRCs, we detected an increase in DAF-FM fluorescence in the anterior neurosecretory neuropil, the region of INNOS projections. This increase did not occur in larvae where we illuminated a control area  **(Figure 2)**.

@@

We next tested whether NO signalling is required for UV/violet avoidance.
To achieve this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method.
We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to frame-shift and an early stop codon and thus likely representing null alleles **(Figure 3---figure supplement 1A)**.
We could establish a homozygous line for both mutations indicating that NOS is not an essential gene in *Platynereis*.
To quantify UV avoidance, we recorded the trajectories of freely swimming wild type and mutant larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm UV light **(Figure 3A and Figure 3---figure supplement 1B)**.
As previously shown, wild-type larvae swim downward following non-directional UV/violet light stimulation [@Veraszto2018].
In contrast, both two- and three-day-old homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response **(Figure 3A, B and Figure 3---figure supplement 1B,C)**.
This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour [@Veraszto2018].
Wild type but not mutant larvae also showed an increase in swimming speed under UV light that may be due to downward swimming trajectories (swimming in the direction of gravity) **(Figure 3B and Figure 3---figure supplement 1C)**.
We also tested directional phototaxis, by exposing larvae to 480 nm directional collimated light from the top of the column.
Three-day-old but not two-day-old *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function for *NOS* in the visual eyes that mediate three-day-old phototaxis [@Randel2014a] **(Figure 3D and Figure 3---figure supplement 1G)**.
To distinguish between an acute and developmental function of NOS in light responses, we next tested larvae exposed to the NOS inhibitor L-NAME.
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a dose-dependent inhibition of UV avoidance.
In contrast, phototaxis was not affected **(Figure 3C, E)**.
Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system, reminiscent of the function of NO signalling in *Drosophila* eye development [@Gibbs1998a].

@@

To investigate how NO signalling alters the dynamics of the cPRC circuit, we carried out calcium imaging experiments.
We ubiquitously expressed the calcium sensor GCaMP6s in larvae and imaged calcium signals during 405 nm stimulation of the cPRCs.
As we have shown previously, a 20-sec local stimulation of cPRC cilia lead to a transient increase in cPRC calcium levels, followed by a transient decrease [@Veraszto2018].
After \~20-sec, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus -- a response that may involve depolarisation **(Figure 4A, B)**.
This activation phase occurs after the 20 sec stimulation period and is likely due to a delayed neuroendocrine feedback [@Veraszto2018].
To determine whether NO mediates such a feedback, we repeated the experiment in *NOS*-mutant larvae.
While we detected the initial activation phase followed by inhibition, in homozygous *NOS*-mutants for both CRISPR alleles this was not followed by delayed activation. Instead, calcium levels dropped to a low steady-state level **(Figure 4A, B)**.
We thus identified a requirement for NO signalling in the late-phase activation of cPRCs.

@@

We aimed next to identify the NO-receptor in the cPRCs.
NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PFAM domain: PF00211).
NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production.
Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit [@Veraszto2017].
Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020].
To identify NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs **(Figure 4---figure supplement 1 and 2)**.
To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup.
In agreement with Moroz et al. [@Moroz2020], we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans.
The 15 *Platynereis* sequences belonged to several deeply diverged clades in the phylogenetic tree **(Figure 4---figure supplement 1 and 2)**.
To characterise the expression of NIT-GCs, we used previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017].
Among the 15 NIT-GCs, two showed high and specific expression in the cPRCs and one was expressed in the INNOS cells **(Figure 4---figure supplement 2)**.
In the single-cell data, we could identify the cPRCs by the specific expression of *c-opsin1* and the *pedal-peptide2 neuropeptide precursor* (*MLD proneuropeptide*), previously described cPRC markers [@Arendt2004; @williams2017] **(Figure 4---figure supplement 3A)**.
The INNOS cells were identified by NOS expression and spatial mapping in the brain [@Achim2015].
We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequence, *NIT-GC1* and *NIT-GC2*.
To confirm the single-cell data, we first carried out in situ hybridisation chain reaction (HCR) with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes were specifically expression in the four cPRCs, as confirmed by co-labeling with an acetylated α-tubulin antibody and with an HCR probe against *pedal peptide 2/MLD proneuropeptide* **(Figure 4C, D and Figure 4---figure supplement 3A-C)**.
To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins.
In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus **(Figure 4E)**.
Co-immunostaining with the rabbit NIT-GC1 antibody and a custom rat antibody raised against *Platynereis* NOS revealed the localisation of both proteins in close proximity in the neurosecretory plexus **(Figure 4---figure supplement 3D)**.
In contrast, NIT-GC2 specifically labelled the ramified sensory cilia of the cPRCs **(Figure 4F)**.
These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

To further characterise these atypical guanylyl cyclases, we focused on NIT-GC1 and carried out in vitro experiments.
In bacteria, NIT domains are thought to regulate cellular functions in response to intra- or extracellular nitrate and nitrite.
NIT-GC1 has a NIT domain and a highly conserved cyclase domain that is expected to catalyse cGMP synthesis **(Figure 4G)**.
The NIT domain may render NIT-GC1 dependent on NO signals.
To test this, we co-expressed the cGMP indicator Green cGull [@Matsuda2017] and NIT-GC1 in cultured COS-7 (monkey kidney) cells, a cell line with minimal endogenous sGC activity.
For balanced expression, we used a single plasmid with the two open-reading frames separated by the 2A self-cleaving peptide **(Figure 4H)**.
Application of the NO donor SNAP lead to increased cGMP levels, an effect we did not observe when cells were exposed to DMSO or when Green cGull was expressed alone **(Figure 4I-K)**.
To test whether cGMP production is dependent on the NIT domain, we also tested a deletion construct of NIT-GC1 lacking the NIT domain **(Figure 4G)**.
Cells expressing this construct and Green cGull did not show increased cGMP levels when exposed to SNAP **(Figure 4L)**.
These results indicate that NIT-GC1 is able to catalyse cGMP production in an NO-dependent manner and this function requires the NIT domain.
These results establish NIT-GC1 as a biochemical NO sensor.

@@

To test the in vivo function of NIT-GC1 and NIT-GC2 in cPRC responses, we combined calcium imaging with morpholino-mediated knockdowns.
We used two translation-blocking morpholinos for each *NIT-GC* gene and tested knockdown efficiency by immunostaining injected animals with the NIT-GC1 and NIT-GC2 antibodies **(Figure 4---figure supplement 3E,F)**.
For both genes, the morpholinos led to a strong reduction in the respective antibody signal, confirming efficient knockdown and antibody specificity.

In NIT-GC1 morphant larvae, the delayed activation of cPRCs following 405 nm stimulation did not occur **(Figure 4M)**.
This phenotype is similar to the phenotype of *NOS* mutants suggesting that NIT-GC1 acts as the NO sensor in cPRCs to drive their delayed activation.
This could occur via increased cGMP production and the opening of a cyclic-nucleotide-gated (CNG) channel specific to cPRCs [@Tosches2014].
NIT-GC2 morphant larvae, in contrast, showed a step-up increase in calcium following light stimulation **(Figure 4N)**.
The calcium signal decayed during stimulation and was off after light off.
These data support an essential role for ciliary-localised NIT-GC2 in suppressing cPRC calcium following its transient rise at stimulus onset.
Overall, these knockdown experiments revealed different signalling mechanisms for the two NIT-GCs that may be due to their different subcellular localisations.

@@

To investigate how NO and NIT-GC signalling influence the dynamics of the cPRC circuit, we imaged calcium signals from postsynaptic neurons in wild type, mutant and morphant larvae.
We were able to image the activity of all neurons in the cPRC circuit (INNOS, INRGW, Ser-h1 and MC).
The MC cell was identified based on its position and intrinsic activity [@Veraszto2017].
To unambiguously identify all other cells from which we recorded calcium signals, we developed an on-slide immunostaining method **(Figure 5---figure supplement 1A)**.
We used the cell-specific antibody markers against RYamide (INNOS) **(Figure 5---figure supplement 1B-D)**, RGWamide (INRGW) and serotonin (Ser-h1) [@Conzelmann2011] to immunostain agar-embedded larvae following calcium imaging.
Based on the position of the nuclei, we could correlate live and fixed samples at a single-cell precision **(Figure 5A,B)**.
Due to the stereotypy of the larvae, we could also identify neurons based on their position and calcium activity in activity-correlation maps **(Figure 5C)**.

We first quantified the responses of the INNOS and INRGW interneurons during 405 nm stimulation of the cPRCs.
In both wild type and *NOS*-mutant larvae, INNOS cells showed an increase in calcium during stimulation **(Figure 5D)**.
In contrast, the INNOS response was flat or slightly negative in NIT-GC2 morphant larvae **(Figure 5E)** revealing an essential role for NIT-GC2-mediated cPRC suppression in INNOS activation.
INRGW cells were initially inhibited during cPRC stimulation, followed by a delayed activation paralleling the second activation phase of cPRCs.
This late INRGW response was lacking in *NOS*-mutants **(Figure 5F)**.
In NIT-GC2 morphants, INRGW cells showed a transient increase in calcium that decayed after light off and a delayed activation was not present **(Figure 5G)**.
Next, we imaged calcium signals from Ser-h1 and MC neurons in wild type and *NOS* mutant larvae.
Ser-h1 cells showed an activation profile that correlated with cPRC activity, including a reduction in calcium during stimulation followed by rebound, a response that was defective in *NOS* mutants **(Figure 5I)**.
MC cells showed sustained activation, including a late-phase that was lacking in *NOS* mutants **(Figure 5I)**.
These data suggest that during 405 nm stimulation the Ser-h1 cells are inhibited and MC cells are activated, and this regulation is NO-dependent.
This pattern is expected to inhibit ciliary activity in the prototroch but not in the other ciliary bands, triggering NO-dependent downward swimming.

@@

We analysed the neurotransmitter-related genes expressed in the respective cells identified from the single-cell transcriptome data of Achim et al. to understand the neurotransmission mechanisms in cPRC, INNOS and INRGWa [@Achim2015; @williams2017].
Interestingly, in cPRC, whereas cholinergic neurons has previously been implicated in neurotransmission [@Randel2014a], we found transporters and synthetic enzymes involved in glutaminergic, glycinergic, GABAergic and adrenergic innervation, respectively **(Figure 6A)**.
Gene expression has been observed by Randel et al. in these genes, suggesting that they may be expressed in cPRC [@Randel2014a].
Their receptors were also found for each of the expressed genes within INNOS and INRGWa **(Figure 6A)**.
Based on these expression patterns and results, a model of transduction across the cPRC pathway focusing on Ca2+, cGMP and NO dynamics is proposed **(Figure 6B)**.
The c-opsin1, which is UV-sensitive, activates G proteins (Gi/Go) and decreases cGMP CNGAα, which is already known to be strongly expressed in cPRC as one of the targets highly dependent on cGMP changes, may regulate the influx of ions into the cell [@Arendt2004; @tsukamoto2017; @Veedin2021].
The Gβγ released by Gi/o activation also induces K+ efflux (hyperpolarisation) via GIRK channels and closes the voltage-gated Ca2+ channels, thereby reducing intracellular calcium levels[@tsukamoto2023].
It is also thought that increased calcium levels in INNOS signalled from cPRC may activate NOS and generate nitric oxide; INNOS has glycinergic and glutamatergic potential and may act as an inhibitory signal regulator for INRGWa.
INRGWa, and may function as an inhibitory signal modulator for INRGWa.

>>>>

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences (Figure 1---figure supplement 1).
To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis.
In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region (Figure 1D and Figure 1---figure supplement 2).
*NOS* was also expressed in the region of the visual eyes (adult eyes) and the pigmented eyespots (Figure 1---figure supplement 2).
The four apical organ cells, but not the eyes, were also labelled with a *NOS* reporter construct driving palmitoylated-tdTomato (Figure 1E).
This reporter also revealed the axonal projections of these central NOS-expressing neurons.
The position and morphology of the four *NOS+* cells allowed us to identify the same four cells as four interneurons (INNOS) in our three-day-old whole-body *Platynereis* volume EM data [@williams2017; @veraszto2020] (Figure 1B,C).
In the synaptic connectome, the INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Figure 1C,F). The projections of INNOS cells are segregated into input and output compartments, with cPRCs inputs occurring in the dense neurosecretory plexus. INNOS output synapses form in the more ventral projection region (Figure 1---figure supplement 3).
INRGW neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the prototroch ciliary band and the cholinergic MC ciliomotor neurons (Figure 1C,F) [@Veraszto2017].

@@

The expression of *NOS* in the INNOS interneurons in the cPRC circuit suggests that NO signalling may be involved in UV/violet-avoidance.
To test this, first we asked whether NO is produced during UV/violet stimulation of the larvae.
We injected the fluorescent NO-reporter DAF-FM into zygotes and imaged two-day-old larvae while exposing the region of cPRC cilia to 405 nm violet light. To mark cell outlines, we coinjected mRNA encoding a red-fluorescent reporter (RGECO), allowing the identification of the cPRCs.
Following light stimulation in the region of the ramified cilia of the cPRCs, we detected an increase in DAF-FM fluorescence in the anterior neurosecretory neuropil, the region of INNOS projections. This increase did not occur in larvae where we illuminated a control area  (Figure 2).

@@

We next tested whether NO signalling is required for UV/violet avoidance.
To achieve this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method.
We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to frame-shift and an early stop codon and thus likely representing null alleles (Figure 3---figure supplement 1A).
We could establish a homozygous line for both mutations indicating that NOS is not an essential gene in *Platynereis*.
To quantify UV avoidance, we recorded the trajectories of freely swimming wild type and mutant larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm UV light (Figure 3A and Figure 3---figure supplement 1B).
As previously shown, wild-type larvae swim downward following non-directional UV/violet light stimulation [@Veraszto2018].
In contrast, both two- and three-day-old homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 3A, B and Figure 3---figure supplement 1B,C).
This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour [@Veraszto2018].
Wild type but not mutant larvae also showed an increase in swimming speed under UV light that may be due to downward swimming trajectories (swimming in the direction of gravity) (Figure 3B and Figure 3---figure supplement 1C).
We also tested directional phototaxis, by exposing larvae to 480 nm directional collimated light from the top of the column.
Three-day-old but not two-day-old *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function for *NOS* in the visual eyes that mediate three-day-old phototaxis [@Randel2014a] (Figure 3D and Figure 3---figure supplement 1G).
To distinguish between an acute and developmental function of NOS in light responses, we next tested larvae exposed to the NOS inhibitor L-NAME.
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a dose-dependent inhibition of UV avoidance.
In contrast, phototaxis was not affected (Figure 3C, E).
Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system, reminiscent of the function of NO signalling in *Drosophila* eye development [@Gibbs1998a].

@@

To investigate how NO signalling alters the dynamics of the cPRC circuit, we carried out calcium imaging experiments.
We ubiquitously expressed the calcium sensor GCaMP6s in larvae and imaged calcium signals during 405 nm stimulation of the cPRCs.
As we have shown previously, a 20-sec local stimulation of cPRC cilia lead to a transient increase in cPRC calcium levels, followed by a transient decrease [@Veraszto2018].
After \~20-sec, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus -- a response that may involve depolarisation (Figure 4A, B).
This activation phase occurs after the 20 sec stimulation period and is likely due to a delayed neuroendocrine feedback [@Veraszto2018].
To determine whether NO mediates such a feedback, we repeated the experiment in *NOS*-mutant larvae.
While we detected the initial activation phase followed by inhibition, in homozygous *NOS*-mutants for both CRISPR alleles this was not followed by delayed activation. Instead, calcium levels dropped to a low steady-state level (Figure 4A, B).
We thus identified a requirement for NO signalling in the late-phase activation of cPRCs.

@@

We aimed next to identify the NO-receptor in the cPRCs.
NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PFAM domain: PF00211).
NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production.
Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit [@Veraszto2017].
Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020].
To identify NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs (Figure 4---figure supplement 1 and 2).
To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan and 2 choanoflagellate species.
We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup.
In agreement with Moroz et al. [@Moroz2020], we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans.
The 15 *Platynereis* sequences belonged to several deeply diverged clades in the phylogenetic tree (Figure 4---figure supplement 1 and 2).
To characterise the expression of NIT-GCs, we used previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017].
Among the 15 NIT-GCs, two showed high and specific expression in the cPRCs and one was expressed in the INNOS cells (Figure 4---figure supplement 2).
In the single-cell data, we could identify the cPRCs by the specific expression of *c-opsin1* and the *pedal-peptide2 neuropeptide precursor* (*MLD proneuropeptide*), previously described cPRC markers [@Arendt2004; @williams2017] (Figure 4---figure supplement 3A).
The INNOS cells were identified by NOS expression and spatial mapping in the brain [@Achim2015].
We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequence, *NIT-GC1* and *NIT-GC2*.
To confirm the single-cell data, we first carried out in situ hybridisation chain reaction (HCR) with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes were specifically expression in the four cPRCs, as confirmed by co-labeling with an acetylated α-tubulin antibody and with an HCR probe against *pedal peptide 2/MLD proneuropeptide* (Figure 4C, D and Figure 4---figure supplement 3A-C).
To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins.
In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus (Figure 4E).
Co-immunostaining with the rabbit NIT-GC1 antibody and a custom rat antibody raised against *Platynereis* NOS revealed the localisation of both proteins in close proximity in the neurosecretory plexus (Figure 4---figure supplement 3D).
In contrast, NIT-GC2 specifically labelled the ramified sensory cilia of the cPRCs (Figure 4F).
These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

To further characterise these atypical guanylyl cyclases, we focused on NIT-GC1 and carried out in vitro experiments.
In bacteria, NIT domains are thought to regulate cellular functions in response to intra- or extracellular nitrate and nitrite.
NIT-GC1 has a NIT domain and a highly conserved cyclase domain that is expected to catalyse cGMP synthesis (Figure 4G).
The NIT domain may render NIT-GC1 dependent on NO signals.
To test this, we co-expressed the cGMP indicator Green cGull [@Matsuda2017] and NIT-GC1 in cultured COS-7 (monkey kidney) cells, a cell line with minimal endogenous sGC activity.
For balanced expression, we used a single plasmid with the two open-reading frames separated by the 2A self-cleaving peptide (Figure 4H).
Application of the NO donor SNAP lead to increased cGMP levels, an effect we did not observe when cells were exposed to DMSO or when Green cGull was expressed alone (Figure 4I-K).
To test whether cGMP production is dependent on the NIT domain, we also tested a deletion construct of NIT-GC1 lacking the NIT domain (Figure 4G).
Cells expressing this construct and Green cGull did not show increased cGMP levels when exposed to SNAP (Figure 4L).
These results indicate that NIT-GC1 is able to catalyse cGMP production in an NO-dependent manner and this function requires the NIT domain.
These results establish NIT-GC1 as a biochemical NO sensor.

@@

To test the in vivo function of NIT-GC1 and NIT-GC2 in cPRC responses, we combined calcium imaging with morpholino-mediated knockdowns.
We used two translation-blocking morpholinos for each *NIT-GC* gene and tested knockdown efficiency by immunostaining injected animals with the NIT-GC1 and NIT-GC2 antibodies (Figure 4---figure supplement 3E,F).
For both genes, the morpholinos led to a strong reduction in the respective antibody signal, confirming efficient knockdown and antibody specificity.

In NIT-GC1 morphant larvae, the delayed activation of cPRCs following 405 nm stimulation did not occur (Figure 4M).
This phenotype is similar to the phenotype of *NOS* mutants suggesting that NIT-GC1 acts as the NO sensor in cPRCs to drive their delayed activation.
This could occur via increased cGMP production and the opening of a cyclic-nucleotide-gated (CNG) channel specific to cPRCs [@Tosches2014].
NIT-GC2 morphant larvae, in contrast, showed a step-up increase in calcium following light stimulation (Figure 4N).
The calcium signal decayed during stimulation and was off after light off.
These data support an essential role for ciliary-localised NIT-GC2 in suppressing cPRC calcium following its transient rise at stimulus onset.
Overall, these knockdown experiments revealed different signalling mechanisms for the two NIT-GCs that may be due to their different subcellular localisations.

@@

To investigate how NO and NIT-GC signalling influence the dynamics of the cPRC circuit, we imaged calcium signals from postsynaptic neurons in wild type, mutant and morphant larvae.
We were able to image the activity of all neurons in the cPRC circuit (INNOS, INRGW, Ser-h1 and MC).
The MC cell was identified based on its position and intrinsic activity [@Veraszto2017].
To unambiguously identify all other cells from which we recorded calcium signals, we developed an on-slide immunostaining method (Figure 5---figure supplement 1A).
We used the cell-specific antibody markers against RYamide (INNOS) (Figure 5---figure supplement 1B-D), RGWamide (INRGW) and serotonin (Ser-h1) [@Conzelmann2011] to immunostain agar-embedded larvae following calcium imaging.
Based on the position of the nuclei, we could correlate live and fixed samples at a single-cell precision (Figure 5A,B).
Due to the stereotypy of the larvae, we could also identify neurons based on their position and calcium activity in activity-correlation maps (Figure 5C).

We first quantified the responses of the INNOS and INRGW interneurons during 405 nm stimulation of the cPRCs.
In both wild type and *NOS*-mutant larvae, INNOS cells showed an increase in calcium during stimulation (Figure 5D).
In contrast, the INNOS response was flat or slightly negative in NIT-GC2 morphant larvae (Figure 5E) revealing an essential role for NIT-GC2-mediated cPRC suppression in INNOS activation.
INRGW cells were initially inhibited during cPRC stimulation, followed by a delayed activation paralleling the second activation phase of cPRCs.
This late INRGW response was lacking in *NOS*-mutants (Figure 5F).
In NIT-GC2 morphants, INRGW cells showed a transient increase in calcium that decayed after light off and a delayed activation was not present (Figure 5G).
Next, we imaged calcium signals from Ser-h1 and MC neurons in wild type and *NOS* mutant larvae.
Ser-h1 cells showed an activation profile that correlated with cPRC activity, including a reduction in calcium during stimulation followed by rebound, a response that was defective in *NOS* mutants (Figure 5I).
MC cells showed sustained activation, including a late-phase that was lacking in *NOS* mutants (Figure 5I).
These data suggest that during 405 nm stimulation the Ser-h1 cells are inhibited and MC cells are activated, and this regulation is NO-dependent.
This pattern is expected to inhibit ciliary activity in the prototroch but not in the other ciliary bands, triggering NO-dependent downward swimming.

@@

We analysed the neurotransmitter-related genes expressed in the respective cells identified from the single-cell transcriptome data of Achim et al. to understand the neurotransmission mechanisms in cPRC, INNOS and INRGWa [@Achim2015; @williams2017].
Interestingly, in cPRC, whereas cholinergic neurons has previously been implicated in neurotransmission [@Randel2014a], we found transporters and synthetic enzymes involved in glutaminergic, glycinergic, GABAergic and adrenergic innervation, respectively (Figure 6A).
Gene expression has been observed by Randel et al. in these genes, suggesting that they may be expressed in cPRC [@Randel2014a].
Their receptors were also found for each of the expressed genes within INNOS and INRGWa (Figure 6A).
Based on these expression patterns and results, a model of transduction across the cPRC pathway focusing on Ca2+, cGMP and NO dynamics is proposed (Figure 6B).
The c-opsin1, which is UV-sensitive, activates G proteins (Gi/Go) and decreases cGMP CNGAα, which is already known to be strongly expressed in cPRC as one of the targets highly dependent on cGMP changes, may regulate the influx of ions into the cell [@Arendt2004; @tsukamoto2017; @Veedin2021].
The Gβγ released by Gi/o activation also induces K+ efflux (hyperpolarisation) via GIRK channels and closes the voltage-gated Ca2+ channels, thereby reducing intracellular calcium levels[@tsukamoto2023].
It is also thought that increased calcium levels in INNOS signalled from cPRC may activate NOS and generate nitric oxide; INNOS has glycinergic and glutamatergic potential and may act as an inhibitory signal regulator for INRGWa.
INRGWa, and may function as an inhibitory signal modulator for INRGWa.
